TIME STRETCHING OF THE GEV EMISSION OF GRBS: FERMI LAT DATA VS GEOMETRICAL MODEL

### 2014-05-25 ITEPProceedings/piskunov.tex
Revised Model section of ITEP Proceedings

	Second, consider the plasma right after its ejection from the central engine.
	Two processes happen there simultaneously:
	\begin{enumerate}
		\item{Particles near the jet boundary change their movement directions, therefore increasing the opening angle of the jet.}
		\item{Particles lose energy, therefore decreasing radiation frequency emitted by the jet.}
	\end{enumerate}
	We argue that these two processes are correlated, for they happen due to the same particle interaction processes.
	And since they are correlated, jets with larger opening angles should have lower energies, which is the assumption we are trying to justify.

	You can see why this assumption allows for non-trivial stretching factors in the figure \ref{fig:modelOverview}.
	More rigorously, the model is defined by the following assumptions:
	\begin{enumerate}
		\item{
			Time $t = 0$, a spherical shell of plasma is emitted.
			The center is called the central engine.
		}
		\item{The shell points propagate with a constant velocity $v = \frac{\sqrt{\gamma^2-1}}{\gamma} \sim 1$, so at the time $t$ the radius of the shall is $v t$.}
		\item{Each point of the shell is an isotropic radiator in its rest frame.}
		\item{
			The radiation intensity is a function of the radiator position and the radiation frequency:
			\begin{equation*}
				\eta\left(r,\theta,\omega\right) = 
				\frac{\eta_0}{1 + \left(\frac{r}{r_0}\right)^n}
				\exp\left(
					-\left(\frac{\theta}{\theta_0}\right)^2
					\left(\frac{\omega}{\omega_0}\right)^{-2k}
				\right)
				\left(\frac{\omega}{\omega_0}\right)^\alpha
			\end{equation*}
			$\eta$ is a number of particles emitted per volume per solid angle per frequency.
			It is a function of the distance $r$ from the central engine, of the off-axis angle $\theta$, and of the radiation frequency $\omega$.
		}
	\end{enumerate}

	The burst is fully specified by the following set of parameters (we assume these parameters to be the same for all bursts, the values are obtained by fitting to make possible observed bursts from section \ref{sec:data}):
	\begin{itemize}
		\item{$\gamma = 300$, the relativistic factor of the shell, $\gamma \gg 1$.}
		\item{$\eta_0 = \unit[8.15 \times 10^{34}]{sec^{-3} GeV^{-1}}$, which defines the luminosity scale.}
		\item{$r_0 = \unit[3.36 \times 10^6]{sec^3}$, the characteristic jet length; $r_0 \ll \frac{1}{H\left(0\right)}$, $H\left(t\right)$ is the Hubble parameter;}
		\item{$n = 22.5$, which determines the sharpness of the jet end, $n > 3$;}
		\item{$\omega_0 = \unit[1]{GeV}$, a characteristic radiation frequency;}
		\item{$\theta_0 = 5.02 \times 10^{-5}$, the opening angle of the jet for radiation with frequency $\omega_0$, $\theta_0 \ll 1$;}
		\item{$k = -0.700$, which determines how much the opening angle changes with frequency, $k < 0$;}
		\item{$\alpha = -0.627$, the bare spectral index, $\alpha < -2k - 1$}
	\end{itemize}

	Numerical simulations based on these assumptions are able to reproduce both low (like of GRB 090902B) and high (like of GRB 090926A) stretching factors (see fig. \ref{fig:sampleLightCurves}).

	To test the model agreement to observations, we computed a few more things in its framework:
	\begin{itemize}
		\item{
			Total energy emitted in $\unit[100]{MeV}$ and more energetic gamma rays $E < \unit[5.89 \times 10^{53}]{GeV}$, which is in agreement with \cite{Gehrels:2013xd}.
		}
		\item{
			Distribution of stretching factors of observable bursts.
			You can see the PDF and CDF on fig. \ref{fig:kappaDistribution}.
			This distribution doesn't contradict with the values of stretching factors computed in section \ref{sec:data}.
		}
		\item{
			The fraction of bursts observable in low energy band, which are also observed in high energy band $f_m = 0.116$.
			The Fermi LAT catalog contains $35$ bursts out of which $4$ can be observed in high energy band.
			Therefore the observed value $f_o = 0.11 \pm 0.05$ (error obtained from binomial distribution) agrees with the model.
		}
		\item{
			There is a correlation between the fractions of high energy photons (that is high energy photon count divided by total photon count), stretching factors and off-axis angles of observers (see fig. \ref{fig:correlations}). It allows one to predict off-axis angles.
		}
	\end{itemize}

@@

	There, however, directions for improvement:
	\begin{enumerate}
		\item{
			The model prediction and observations on fig. \ref{fig:correlations} are, though close, not exactly agree.
		}
		\item{
			Shapes of light curves produced by model differ considerably from observed ones.
		}
		\item{
			Assumption of energy dependence on off-axis angle should be better justified.
		}
		\item{
			Much more results can be obtained from the model (like energy spectrum, and its variation with time).
		}
	\end{enumerate}

>>>>

	Second, consider the plasma right after its ejection from the central engine.
	Two processes happen here simultaneously:
	\begin{enumerate}
		\item{Particles near the jet boundary change their movement directions, therefore increasing the opening angle of the jet.}
		\item{Particles lose energy, therefore decreasing radiation frequency emitted by the jet.}
	\end{enumerate}
	We argue that these two processes are correlated, for they happen due to the same particle interaction processes.
	And since they are correlated, jets with larger opening angles should have lower energies, which is the assumption we are trying to justify.

	You can see why this assumption allows for non-trivial stretching factors in the figure \ref{fig:modelOverview}.
	More rigorously, the model is defined by the following assumptions:
	\begin{enumerate}
		\item{
			Time $t = 0$, a spherical shell of plasma is emitted.
			The center is called the central engine.
		}
		\item{The shell points propagate with a constant velocity $v = \frac{\sqrt{\gamma^2-1}}{\gamma} \sim 1$, so at the time $t$ the radius of the shall is $v t$.}
		\item{Each point of the shell is an isotropic radiator in its rest frame.}
		\item{
			The radiation intensity is a function of the radiator position and the radiation frequency:
			\begin{equation*}
				\eta\left(r,\theta,\omega\right) = 
				\frac{\eta_0}{1 + \left(\frac{r}{r_0}\right)^n}
				\exp\left(
					-\left(\frac{\theta}{\theta_0}\right)^2
					\left(\frac{\omega}{\omega_0}\right)^{-2k}
				\right)
				\left(\frac{\omega}{\omega_0}\right)^\alpha
			\end{equation*}
			$\eta$ is the number of particles emitted per volume per solid angle per frequency.
			It is a function of the distance $r$ from the central engine, of the off-axis angle $\theta$, and of the radiation frequency $\omega$.
		}
	\end{enumerate}

	The burst is fully specified by the following set of parameters (we assume these parameters to be the same for all bursts, the specified values are obtained by fitting to make possible observed bursts from section \ref{sec:data}):
	\begin{itemize}
		\item{$\gamma = 300$, the relativistic factor of the shell, $\gamma \gg 1$.}
		\item{$\eta_0 = \unit[8.15 \times 10^{34}]{sec^{-3} GeV^{-1}}$, which defines the luminosity scale.}
		\item{$r_0 = \unit[3.36 \times 10^6]{sec^3}$, the characteristic jet length; $r_0 \ll \frac{1}{H\left(0\right)}$, $H\left(t\right)$ is the Hubble parameter;}
		\item{$n = 22.5$, which determines the sharpness of the jet end, $n > 3$;}
		\item{$\omega_0 = \unit[1]{GeV}$, a characteristic radiation frequency;}
		\item{$\theta_0 = 5.02 \times 10^{-5}$, the opening angle of the jet for radiation with frequency $\omega_0$, $\theta_0 \ll 1$;}
		\item{$k = -0.700$, which determines how much the opening angle changes with frequency, $k < 0$;}
		\item{$\alpha = -0.627$, the bare spectral index, $\alpha < -2k - 1$}
	\end{itemize}

	Numerical simulations based on these assumptions are able to reproduce both stretching factors smaller than $1$ (like of GRB 090902B) and larger than $1$ (like of GRB 090926A), see fig. \ref{fig:sampleLightCurves}.

	To test the model against observations, we computed a few more things:
	\begin{itemize}
		\item{
			Total energy emitted in $\unit[100]{MeV}$ and more energetic gamma rays $E < \unit[5.89 \times 10^{53}]{GeV}$, which is in agreement with \cite{Gehrels:2013xd}.
		}
		\item{
			Distribution of stretching factors of observable bursts.
			You can see the PDF and the CDF of it on fig. \ref{fig:kappaDistribution}.
			This distribution doesn't contradict with the values of stretching factors obtained in section \ref{sec:data}.
		}
		\item{
			The fraction of bursts observable in low energy band, which are also observed in high energy band $f_m = 0.116$.
			The Fermi LAT catalog contains $35$ bursts out of which $4$ can be observed in high energy band.
			Therefore the observed value $f_o = 0.11 \pm 0.05$ (error obtained from binomial distribution) agrees with the model.
		}
		\item{
			There is a correlation between the fraction of high energy photons (that is high energy photon count divided by the total photon count), the stretching factor and the observer's off-axis angle (see fig. \ref{fig:correlations}). It allows one to predict observer's off-axis angles.
		}
	\end{itemize}

@@

	There, however, directions for improvement:
	\begin{itemize}
		\item{
			The model prediction and observations on fig. \ref{fig:correlations} are, though close, not exactly agree.
		}
		\item{
			Shapes of light curves produced by model differ considerably from observed ones.
		}
		\item{
			Assumption of energy dependence on off-axis angle should be better justified.
		}
		\item{
			Much more results can be obtained from the model (like energy spectrum, and its variation with time).
		}
	\end{itemize}
